Generation of TypeScript Declaration Files from JavaScript Code

### 2020-09-26 programming-2021/main.tex
final changes pre submission

\section{Evaluation}
\label{sec:dts-generate-evaluation}
After generating a declaration file for an \NPM{} package, we need to
evaluate its quality. To this end, we created \texttt{dts-compare}, a tool to compute the
differences between two TypeScript declaration files.
% \begin{itemize}
%   \item \texttt{dts-parse}: A TypeScript declaration files parser, which transforms the file into a JSON file with a defined structure.
%   \item \texttt{dts-compare}: A tool that uses \texttt{dts-parse} and compares two TypeScript declaration files computing differences by applying several criteria.
% \end{itemize}

For each module, we applied \texttt{dts-compare} to the generated declaration file and the
one uploaded to the DefinitelyTyped repository for the same module as shown in
\figref{fig:evaluation-diagram}. While this approach does not provide an absolute measure
of quality, it gives us at least an indication of the pragmatics of \texttt{dts-generate}:
The files on DefinitelyTyped are perceived to be useful by the community. If the accuracy
of the generated files is comparable with the accuracy of the files on DefinitelyTyped,
then \texttt{dts-generate} is a viable alternative to manually created definition files.

@@

\subsection{dts-parse}
\label{sec:dts-parse}
We need a means of comparing two declaration files. Of
course, we are not interested in a textual comparison, but in a
comparison of the structures described by the files.

The first step is to parse the declaration files, which can be
achieved by using the TypeScript Compiler API, a library which traverses the
Abstract Syntax Tree of a TypeScript program
\cite{typescript-compiler-api}. 
The step also performs a sanity check of the generated declaration
files as it rejects files with syntactic or semantics errors.
The output of \texttt{dts-parse} is a structure where declared
{interfaces}, {functions}, {classes}, and
{namespaces} are stored separately. Function arguments are
correctly described, identifying complex types like union types or
callbacks. Optional parameters are also identified. For
{classes}, a distinction between constructors and methods
is made. Declaration files are tagged according to their features,
which is useful for identifying and filtering out declaration files
that contain unimplemented features.

@@

% The tool is written in TypeScript.
The compare tool takes two declaration files as input and returns a
JSON file containing the result of the comparison as shown in
\coderef{code:dts-compare-example}. Following the standard naming convention of unit
testing frameworks, the input files are marked as \lstinline{expected} or
\lstinline{actual}. We use \lstinline{expected} for the DefinitelyTyped
declaration files.

\begin{lstlisting}[
  caption={Example for \texttt{dts-compare} executed on declaration files without differences},
  language=bash,
label=code:dts-compare-example,
  float=tp,captionpos=b
]
$ dts-compare -e expected.d.ts -a actual.d.ts --module-name "my-module"
{
    "module": "my-module",
    "template": "module",
    "differences": [],
    "tags": []
}
\end{lstlisting}

\texttt{dts-compare} applies \texttt{dts-parse} internally to both files to get 
normalized structures that are easy to compare. The result of the comparison is an array
of an abstract entity \texttt{Difference}. 
A difference is \texttt{solvable} if it can be resolved by adding more code
examples. For example, invoking a function with different parameters so that a missing
interface property gets accessed. Otherwise it is  \texttt{unsolvable}.

The concrete evaluated differences are as follows:
\begin{itemize}
  \item \texttt{TemplateDifference}: A difference between the choice of TypeScript
    templates. For example, if the file in DefinitelyTyped is written as a \textbf{module}
    but we generated the file using the \textbf{module-function} template. The comparison
    stops if the templates differ. 
  \item \texttt{ExportAssignmentDifference}: An equality check between the export
    assignments of both files, that is in the expression
    \lstinline{export = XXX}.
  \item \texttt{FunctionMissingDifference}: A function declaration is present in the expected
    file but not in the generated one. This difference applies to functions as well as methods of classes
    and properties of interfaces. 
  \item \texttt{FunctionExtraDifference}: A function is present in the generated declaration file but not in DefinitelyTyped.
  \item \texttt{FunctionOverloadingDifference}: The number of declarations for the same
    function is different. This difference can be due to inexperience of the author of a
    declaration file, who uses, e.g., multiple declarations where a union typed argument
    would be appropriate.
  \item \texttt{ParameterMissingDifference}: A parameter of a function or a property of an interface is not present in the generated file.
  \item \texttt{ParameterExtraDifference}: A parameter of a function or a property of an
    interface is present in the generated file but not in the DefinitelyTyped file. 
  \item \texttt{ParameterTypeDifference}: A parameter of a function or a property of an interface is generated with a different type than in the DefinitelyTyped file. Here we differentiate between \texttt{SolvableDifference} and \texttt{UnsolvableDifference}.
  \begin{itemize}
    \item \texttt{SolvableDifference}: A type difference that can be solved by writing
      further code examples. For example, a basic type that is converted to a union type,
      a function overloading, or a parameter or property that is marked as optional. 
    \item \texttt{UnsolvableDifference}: Any difference not considered as \texttt{SolvableDifference}.
  \end{itemize}
\end{itemize}

@@

\begin{changethis}

To measure the impact of incomplete code examples on the
quality of generated declaration files we categorized each difference as follows:

  We extracted code examples written by developers to execute the code
  and gather run-time information. The quality of the generated
  declaration file is heavily dependent on the code examples. For
  example, if a property of an interface does not get accessed due to
  a code example not exploring a particular execution path, that
  property will not exist in the generated declaration file.
\end{changethis}

>>>>

\section{Evaluation}
\label{sec:dts-generate-evaluation}
After generating a declaration file for an \NPM{} package, we need to
evaluate its quality.
As we extracted code examples written by developers, the quality of the generated
declaration file is heavily dependent on these examples. For
example, if a property of an interface is not accessed because
a code example does not explore a particular execution path, then that
property will not appear in the generated declaration file.

As a measure of the quality of a generated file we use its distance to
the declaration file provided on DefinitelyTyped for the same module. 
To this end, we created \texttt{dts-compare}, a tool to compute the
differences between two TypeScript declaration files.
% \begin{itemize}
%   \item \texttt{dts-parse}: A TypeScript declaration files parser, which transforms the file into a JSON file with a defined structure.
%   \item \texttt{dts-compare}: A tool that uses \texttt{dts-parse} and compares two TypeScript declaration files computing differences by applying several criteria.
% \end{itemize}

For each module, we applied \texttt{dts-compare} to the generated declaration file and the
one uploaded to the DefinitelyTyped repository for the same module as shown in
\figref{fig:evaluation-diagram}. While this approach does not provide an absolute measure
of quality, it gives us at least an indication of the pragmatics of \texttt{dts-generate}:
The files on DefinitelyTyped are perceived to be useful by the community. If the accuracy
of the generated files is comparable with the accuracy of the files on DefinitelyTyped,
then \texttt{dts-generate} is a viable alternative to hand-crafted definition files.

@@

\subsection{dts-parse}
\label{sec:dts-parse}

We are not interested in a textual comparison of declaration files, but in a
comparison of the structures described by the files. Hence,
the first step is to parse the declaration files using the TypeScript
Compiler API to build and traverse the 
Abstract Syntax Tree of a TypeScript program
\cite{typescript-compiler-api}. 
This step also performs a sanity check of the generated declaration
files as it rejects files with syntactic or semantic errors.
The output of \texttt{dts-parse} is a structure similar to a symbol table where declared
{interfaces}, {functions}, {classes}, and
{namespaces} are stored separately. Function arguments are
correctly described, identifying complex types like union types or
callbacks. Optional parameters are also identified. For
{classes}, a distinction between constructors and methods
is made. Declaration files are tagged according to their features,
which is useful for identifying and filtering out declaration files
that contain unimplemented features.

@@

% The tool is written in TypeScript.
The compare tool takes two declaration files as input and returns a
JSON file containing the result of the comparison as shown in
\coderef{code:dts-compare-example}. Following the naming convention of unit
testing frameworks, the input files are marked as \lstinline{expected} or
\lstinline{actual}. We use \lstinline{expected} for the DefinitelyTyped
declaration files.

\begin{lstlisting}[
  caption={Example for \texttt{dts-compare} executed on declaration files without differences},
  language=bash,
label=code:dts-compare-example,
  float=tp,captionpos=b
]
$ dts-compare -e expected.d.ts -a actual.d.ts --module-name "my-module"
{
    "module": "my-module",
    "template": "module",
    "differences": [],
    "tags": []
}
\end{lstlisting}
% $

\texttt{dts-compare} applies \texttt{dts-parse} internally to both files to get 
normalized structures that are easy to compare. The result of the comparison is an array
of an abstract entity \texttt{Difference}. 
We classify a difference as \texttt{solvable} if it can be resolved by adding more code
examples. For example, we might add a function invocation with different parameters so that a missing
interface property gets accessed. Otherwise it is classified as  \texttt{unsolvable}.

We consider the following differences:
\begin{itemize}
  \item \texttt{TemplateDifference}: A difference between the choice of TypeScript
    templates. For example, if the file in DefinitelyTyped is written as a \textbf{module}
    but we generated the file using the \textbf{module-function} template. The comparison
    stops if the templates differ. 
  \item \texttt{ExportAssignmentDifference}: An equality check between the export
    assignments of both files, that is in the expression
    \lstinline{export = XXX}.
  \item \texttt{FunctionMissingDifference}: A function declaration is present in the expected
    file but not in the generated one. This difference applies to functions as well as methods of classes
    and properties of interfaces. 
  \item \texttt{FunctionExtraDifference}: A function is present in the generated declaration file but not in DefinitelyTyped.
  \item \texttt{FunctionOverloadingDifference}: The number of declarations for the same
    function is different. This difference can be due to inexperience of the author of a
    declaration file, who uses, e.g., multiple declarations where a union typed argument
    would be appropriate.
  \item \texttt{ParameterMissingDifference}: A parameter of a function or a property of an interface is not present in the generated file.
  \item \texttt{ParameterExtraDifference}: A parameter of a function or a property of an
    interface is present in the generated file but not in the DefinitelyTyped file. 
  \item \texttt{ParameterTypeDifference}: A parameter of a function or a property of an interface is generated with a different type than in the DefinitelyTyped file. Here we differentiate between \texttt{SolvableDifference} and \texttt{UnsolvableDifference}.
  \begin{itemize}
    \item \texttt{SolvableDifference}: A type difference that can be solved by writing
      further code examples. For example, a basic type that is converted to a union type,
      a function overloading, or a parameter or property that is marked as optional. 
    \item \texttt{UnsolvableDifference}: Any difference not considered as \texttt{SolvableDifference}.
  \end{itemize}
\end{itemize}

@@
